Illusion Sensitivity and Phenomenological Control

### 2025-05-21 analysis/1_cleaning.qmd
removed paths for the review

# data_ppt <- read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGamePhenomenologicalControl/refs/heads/main/data/rawdata_participants.csv")

@@

# read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGamePhenomenologicalControl/refs/heads/main/data/rawdata_illusion.csv")
data_ig <- read.csv("../data/rawdata_illusion.csv") |>
  mutate(
    Illusion_Effect = ifelse(sign(Illusion_Strength) > 0, "Incongruent", "Congruent"),
    Task_Difficulty = abs(Illusion_Difference),
    Condition_Illusion = case_when(
        Illusion_Effect == "Congruent" ~ "Congruent",
        Illusion_Effect == "Incongruent" & Illusion_Strength <= median(Illusion_Strength[Illusion_Effect == "Incongruent"], na.rm = TRUE) ~ "Mild",
        Illusion_Effect == "Incongruent" & Illusion_Strength > median(Illusion_Strength[Illusion_Effect == "Incongruent"], na.rm = TRUE) ~ "Strong"),
    Condition_Difficulty = datawizard::categorize(
      Task_Difficulty, 
      split = "quantile", 
      n_groups = 2, 
      labels = c("Hard", "Easy")),
    .by = "Illusion_Type"
  )
```

@@

```{r}
data_ppt |>
  mutate(Participant = fct_reorder(Participant, Experiment_Duration),
         Category = ifelse(Experiment_Duration > 50, "extra", "ok"),
         Duration = ifelse(Experiment_Duration > 50, 50, Experiment_Duration)) |>
  ggplot(aes(y = Participant, x = Duration)) +
  geom_point(aes(color = Category, shape = Category)) +
  geom_vline(xintercept = median(data_ppt$Experiment_Duration), color = "red", linetype = "dashed") +
  scale_shape_manual(values = c("extra" = 3, ok = 19)) +
  scale_color_manual(values = c("extra" = "red", ok = "black")) +
  guides(color = "none", shape = "none") +
  ggside::geom_xsidedensity(fill = "grey", color=NA) +
  ggside::scale_xsidey_continuous(expand = c(0, 0)) +
  labs(
    title = "Experiment Completion Time",
    x = "Duration (in minutes)",
    y = "Participants"
  )  +
  see::theme_modern() +
  ggside::theme_ggside_void() +
  theme(ggside.panel.scale = .3, 
        axis.text.y = element_blank()) 
```

@@

data_outliers <- data_ig |>
  summarize(p_Error = sum(Error) / n(),
            z_RT = mean(RT),
            .by = c("Participant", "Illusion_Type", "Block")) |> 
  mutate(z_RT = as.numeric(standardize(z_RT)), .by = c("Illusion_Type", "Block")) |>
  arrange(desc(p_Error))

@@

p1 <- data_outliers |> 
  ggplot(aes(x = p_Error)) +
  geom_density(aes(color=Illusion_Type)) +
  geom_vline(xintercept = 0.45, linetype = "dashed", color="red")

p2 <- data_outliers |> 
  ggplot(aes(x = z_RT)) +
  geom_density(aes(color=Illusion_Type)) +
  geom_vline(xintercept = 4, linetype = "dashed", color="red")

@@

```{r}
p_age <- data_ppt |>  # To match the binwidth
  ggplot(aes(x = data_ppt)) +
  geom_histogram(aes(x = Age, fill=Gender), bins=28) +
  geom_vline(xintercept = mean(data_ppt$Age), color = "red", linewidth=1.5) +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Age", y = "Number of Participants", color = NULL, subtitle = "Distribution of participants' age") +
  theme_modern(axis.title.space = 10) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_age
```

@@

```{r}
p_edu <- data_ppt |>
  mutate(Education = fct_relevel(Education, "Other", "Elementary school", "High school", "Bachelor", "Master", "Doctorate")) |>
  ggplot(aes(x = Education)) +
  geom_bar(aes(fill = Education)) +
  scale_y_continuous(expand = c(0, 0), breaks= scales::pretty_breaks()) +
  scale_fill_viridis_d(guide = "none") +
  labs(title = "Education", y = "Number of Participants", subtitle = "Participants per achieved education level") +
  theme_modern(axis.title.space = 15) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_edu
```

>>>>

@@

data_ig <- read.csv("../data/rawdata_illusion.csv") |>
    mutate(
        Illusion_Effect = ifelse(sign(Illusion_Strength) > 0, "Incongruent", "Congruent"),
        Task_Difficulty = abs(Illusion_Difference),
        Condition_Illusion = case_when(
            Illusion_Effect == "Congruent" ~ "Congruent",
            Illusion_Effect == "Incongruent" & Illusion_Strength <= median(Illusion_Strength[Illusion_Effect == "Incongruent"], na.rm = TRUE) ~ "Mild",
            Illusion_Effect == "Incongruent" & Illusion_Strength > median(Illusion_Strength[Illusion_Effect == "Incongruent"], na.rm = TRUE) ~ "Strong"
        ),
        Condition_Difficulty = datawizard::categorize(
            Task_Difficulty,
            split = "quantile",
            n_groups = 2,
            labels = c("Hard", "Easy")
        ),
        .by = "Illusion_Type"
    )
```

@@

```{r}
data_ppt |>
    mutate(
        Participant = fct_reorder(Participant, Experiment_Duration),
        Category = ifelse(Experiment_Duration > 50, "extra", "ok"),
        Duration = ifelse(Experiment_Duration > 50, 50, Experiment_Duration)
    ) |>
    ggplot(aes(y = Participant, x = Duration)) +
    geom_point(aes(color = Category, shape = Category)) +
    geom_vline(xintercept = median(data_ppt$Experiment_Duration), color = "red", linetype = "dashed") +
    scale_shape_manual(values = c("extra" = 3, ok = 19)) +
    scale_color_manual(values = c("extra" = "red", ok = "black")) +
    guides(color = "none", shape = "none") +
    ggside::geom_xsidedensity(fill = "grey", color = NA) +
    ggside::scale_xsidey_continuous(expand = c(0, 0)) +
    labs(
        title = "Experiment Completion Time",
        x = "Duration (in minutes)",
        y = "Participants"
    ) +
    see::theme_modern() +
    ggside::theme_ggside_void() +
    theme(
        ggside.panel.scale = .3,
        axis.text.y = element_blank()
    )
```

@@

data_outliers <- data_ig |>
    summarize(
        p_Error = sum(Error) / n(),
        z_RT = mean(RT),
        .by = c("Participant", "Illusion_Type", "Block")
    ) |>
    mutate(z_RT = as.numeric(standardize(z_RT)), .by = c("Illusion_Type", "Block")) |>
    arrange(desc(p_Error))

@@

p1 <- data_outliers |>
    ggplot(aes(x = p_Error)) +
    geom_density(aes(color = Illusion_Type)) +
    geom_vline(xintercept = 0.45, linetype = "dashed", color = "red")

p2 <- data_outliers |>
    ggplot(aes(x = z_RT)) +
    geom_density(aes(color = Illusion_Type)) +
    geom_vline(xintercept = 4, linetype = "dashed", color = "red")

@@

```{r}
p_age <- data_ppt |> # To match the binwidth
    ggplot(aes(x = data_ppt)) +
    geom_histogram(aes(x = Age, fill = Gender), bins = 28) +
    geom_vline(xintercept = mean(data_ppt$Age), color = "red", linewidth = 1.5) +
    scale_fill_manual(values = c("Male" = "#64B5F6", "Female" = "#F06292", "Other" = "orange")) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = "Age", y = "Number of Participants", color = NULL, subtitle = "Distribution of participants' age") +
    theme_modern(axis.title.space = 10) +
    theme(
        plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
        plot.subtitle = element_text(size = rel(1.2), vjust = 7),
        axis.text.y = element_text(size = rel(1.1)),
        axis.text.x = element_text(size = rel(1.1)),
        axis.title.x = element_blank()
    )
p_age
```

@@

```{r}
p_edu <- data_ppt |>
    mutate(Education = fct_relevel(Education, "Other", "Elementary school", "High school", "Bachelor", "Master", "Doctorate")) |>
    ggplot(aes(x = Education)) +
    geom_bar(aes(fill = Education)) +
    scale_y_continuous(expand = c(0, 0), breaks = scales::pretty_breaks()) +
    scale_fill_viridis_d(guide = "none") +
    labs(title = "Education", y = "Number of Participants", subtitle = "Participants per achieved education level") +
    theme_modern(axis.title.space = 15) +
    theme(
        plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
        plot.subtitle = element_text(size = rel(1.2), vjust = 7),
        axis.text.y = element_text(size = rel(1.1)),
        axis.text.x = element_text(size = rel(1.1)),
        axis.title.x = element_blank()
    )
p_edu
```

### 2025-05-21 analysis/2_analysis.qmd
removed paths for the review

```{r}
ig_scores <- data_ig |>
  summarize(ErrorRate = sum(Error) / n(),
            IES = mean(RT[Response_Correct == "true"] / mean(Response_Correct == "true")),
            .by = c("Participant", "Illusion_Type", "Condition_Illusion", "Condition_Difficulty")) |> 
  pivot_wider(names_from = c("Illusion_Type", "Condition_Illusion", "Condition_Difficulty"), 
              values_from = c("ErrorRate", "IES")) 
```

```{r}
corrs <- data.frame()
for(outcome in c("ErrorRate", "IES")) {
  for(ill_type in c("Ebbinghaus", "MullerLyer", "VerticalHorizontal")) {
    for(cond in c("Congruent", "Mild", "Strong")) {
      corrs <- correlation::correlation(
        select(ig_scores, select=paste0(outcome, "_", ill_type, "_", cond, "_Easy")),
        select(ig_scores, select=paste0(outcome, "_", ill_type, "_", cond, "_Hard"))) |> 
        mutate(Outcome = outcome) |> 
        rbind(corrs)
    }
  }
}

# IF correlations > 0.5:
ig_scores <- data_ig |>
  summarize(ErrorRate = sum(Error) / n(),
            IES = mean(RT[Response_Correct == "true"] / mean(Response_Correct == "true")),
            .by = c("Participant", "Illusion_Type", "Condition_Illusion")) |> 
  pivot_wider(names_from = c("Illusion_Type", "Condition_Illusion"), 
              values_from = c("ErrorRate", "IES")) 
```

@@

```{r}
ttests <- data.frame()
for(outcome in c("ErrorRate", "IES")) {
  for(ill_type in c("Ebbinghaus", "MullerLyer", "VerticalHorizontal")) {
    for(cond in c("Congruent", "Mild")) {
      # Add loop for Hard and Easy if need be
      x1 <- ig_scores[[paste0(outcome, "_", ill_type, "_", cond)]]
      if(cond == "Congruent") {
        x2 <- ig_scores[[paste0(outcome, "_", ill_type, "_", "Mild")]]
      } else {
        x2 <- ig_scores[[paste0(outcome, "_", ill_type, "_", "Strong")]]
      }
      ttests <- BayesFactor::ttestBF(x = x1, y = x2, paired = TRUE) |> 
        parameters::parameters() |> 
        mutate(
          Outcome = outcome,
          Illusion_Type = ill_type,
          Condition_Illusion = cond
        ) |> 
        rbind(ttests)
    }
  }
}

>>>>

```{r}
ig_scores <- data_ig |>
    summarize(
        ErrorRate = sum(Error) / n(),
        IES = mean(RT[Response_Correct == "true"] / mean(Response_Correct == "true")),
        .by = c("Participant", "Illusion_Type", "Condition_Illusion", "Condition_Difficulty")
    ) |>
    pivot_wider(
        names_from = c("Illusion_Type", "Condition_Illusion", "Condition_Difficulty"),
        values_from = c("ErrorRate", "IES")
    )
```

```{r}
corrs <- data.frame()
for (outcome in c("ErrorRate", "IES")) {
    for (ill_type in c("Ebbinghaus", "MullerLyer", "VerticalHorizontal")) {
        for (cond in c("Congruent", "Mild", "Strong")) {
            corrs <- correlation::correlation(
                select(ig_scores, select = paste0(outcome, "_", ill_type, "_", cond, "_Easy")),
                select(ig_scores, select = paste0(outcome, "_", ill_type, "_", cond, "_Hard"))
            ) |>
                mutate(Outcome = outcome) |>
                rbind(corrs)
        }
    }
}

# IF correlations > 0.5:
ig_scores <- data_ig |>
    summarize(
        ErrorRate = sum(Error) / n(),
        IES = mean(RT[Response_Correct == "true"] / mean(Response_Correct == "true")),
        .by = c("Participant", "Illusion_Type", "Condition_Illusion")
    ) |>
    pivot_wider(
        names_from = c("Illusion_Type", "Condition_Illusion"),
        values_from = c("ErrorRate", "IES")
    )
```

@@

```{r}
ttests <- data.frame()
for (outcome in c("ErrorRate", "IES")) {
    for (ill_type in c("Ebbinghaus", "MullerLyer", "VerticalHorizontal")) {
        for (cond in c("Congruent", "Mild")) {
            # Add loop for Hard and Easy if need be
            x1 <- ig_scores[[paste0(outcome, "_", ill_type, "_", cond)]]
            if (cond == "Congruent") {
                x2 <- ig_scores[[paste0(outcome, "_", ill_type, "_", "Mild")]]
            } else {
                x2 <- ig_scores[[paste0(outcome, "_", ill_type, "_", "Strong")]]
            }
            ttests <- BayesFactor::ttestBF(x = x1, y = x2, paired = TRUE) |>
                parameters::parameters() |>
                mutate(
                    Outcome = outcome,
                    Illusion_Type = ill_type,
                    Condition_Illusion = cond
                ) |>
                rbind(ttests)
        }
    }
}
